---
title: Why, when and how to latent mean center variables in multilevel models
shorttitle: Latent mean centering
leftheader: Latent mean centering
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Oxford Internet Institute, University of Oxford
    email: matti.vuorre@oii.ox.ac.uk
    role:
      - role mcrole
  - name: Author McAuthor
    affiliation: 2
    address: Address McAddress
    email: email@mcemail.com
    role:
      - role mcrole

affiliation:
  - id: 1
    institution: Oxford Internet Institute, University of Oxford
  - id: 2
    institution: Institution McInstitution

authornote: |
  \noindent \textbf{This manuscript is not yet peer-reviewed.}
  
abstract: |
  Many things, few words.
  
keywords: words, key
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: "references.bib"
floatsintext: yes
linenumbers: yes
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: false
  papaja::apa6_docx: 
    number_sections: false
---

```{r setup, include = FALSE}
library(papaja)

# Parallel computation options
MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
if (is.na(MAX_CORES)) MAX_CORES <- parallel::detectCores(logical = FALSE)
```

# Introduction

## Centering

## Latent mean centering

# Example analysis

# Simulation study

# Discussion

R [@rcoreteamLanguageEnvironmentStatistical2022]

\clearpage

# Acknowledgements

We would like to thank Mauricio Garnier-Villarre, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424). Special thanks to Mauricio Garnier-Villarre for being the first to post the syntax that the code here was based on. We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.

# References

::: {#refs custom-style="Bibliography"}
:::


<!-- Create an appendix with new page, table, and figure numbers. -->

\clearpage
\onecolumn

# Appendix A: Appendix McAppendix

<!-- Prepend table & figure numbers with S and reset counters -->
\setcounter{page}{1}
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r include-only-for-pdf}
#| include: !expr knitr::is_latex_output()
```
