---
title: 'Bias no more: A practical guide to latent group-mean centering in longitudinal and other hierarchical bayesian models'
shorttitle: Latent group-mean centering
leftheader: Latent group-mean centering
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Oxford Internet Institute, University of Oxford
    email: matti.vuorre@oii.ox.ac.uk
    role:
      - role mcrole
  - name: Author McAuthor
    affiliation: 2
    address: Address McAddress
    email: email@mcemail.com
    role:
      - role mcrole

affiliation:
  - id: 1
    institution: Oxford Internet Institute, University of Oxford
  - id: 2
    institution: Institution McInstitution

authornote: |
  \noindent \textit{This manuscript is not yet peer-reviewed.}
  
abstract: |
  Researchers studying the contributions of observed variables on other variables in longitudinal and other repeated measures data sets have long recognized the importance of group-mean centering variables in order to isolate within- and between-person causal processes. However, a critically important issue in centering---treating the person means as unknown parameters rather than observed quantities---has received less attention among applied practitioners. One reason for this has been the general unavailability and perceived complexity of latent group-mean centering versus observed mean centering. 
  In this practical tutorial, we introduce code for latent group-mean centering in longitudinal and hierarchical models using the free and open source software R and specifically the brms R package. We show via analyses of data on well-being and technology use that latent mean centering is not only important, easy, but also fun. We hope that these examples, and the general FOSS availability of the discussed methods encourage more researchers to adopt latent mean centering instead of oserved mean centering to avoid bias in parameter estimates. We also discuss relations between longitudinal models / SEM / multilevel models / bayes. Bayes!
  
keywords: longitudinal data, repeated measures, centering, well-being, technology effects
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: "references.bib"
floatsintext: yes
linenumbers: yes
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: false
  papaja::apa6_docx: 
    number_sections: false
---

```{r setup, include = FALSE}
library(papaja)
library(posterior)
library(scales)
library(ggdist)
library(brms)
library(tidyverse)

# HMC and parallel computation options
MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
if (is.na(MAX_CORES)) MAX_CORES <- parallel::detectCores(logical = FALSE)
THREADS <- 1
CORES <- MAX_CORES
ITER <- 10000
if (require(cmdstanr)) {
  options(brms.backend = "cmdstanr")
  THREADS <- MAX_CORES %/% 4
}

# Save models in a directory
dir.create("models", FALSE)

# Plotting options
theme_set(
  theme_linedraw(base_size = 9.6) +
    theme(
      strip.background = element_rect(fill = "gray90", color = "gray90"),
      strip.text = element_text(color = "black", hjust = 0),
      panel.grid = element_blank()
    )
)
```

# Introduction

## Centering

## Latent mean centering

# Example analysis

<!-- Just a toy example for now to write some code -->

```{r}
# Data from McNeish
d <- read_csv("data/Two-Level Data.csv", col_names = FALSE)[,c(5:6,1:2)]
names(d) <- c("person","time","urge", "dep")
d <- d %>% 
  # Create lagged variable
  group_by(person) %>% 
  mutate(urge1 = lag(urge)) %>% 
  ungroup()

# Model definition from https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424/22
# Estimate depBI (latent means) using nlf() inside bf()
model <- bf(
  urge ~ alpha + phi*(urge1 - alpha) + beta*(dep - depB),
  alpha ~ 1 + (1 | person),
  phi ~ 1 + (1 | person),
  beta ~ 1 + (1 | person),
  nlf(depB ~ depBI),
  depBI ~ 1 + (1 | person),
  nl = TRUE
) +
  gaussian()

p <- get_prior(model, data = d) %>%
  mutate(
    prior = case_when(
      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
      TRUE ~ prior
    )
  )

fit1 <- brm(
  model,
  data = d,
  prior = p,
  cores = MAX_CORES,
  threads = THREADS,
  iter = ITER,
  control = list(adapt_delta = 0.99),
  file = "models/m_mh91"
)

# Code example from Ethan McCormick
# Matti's code edited to add `depb` as 'main effect'
# Estimate model without nlf() with depb added to the first formula
# This seems to be slightly different as it adds depb as predictor of urge?
# But under typical conditions these should give similar key estimates (?)
model <- bf(
  urge ~ alpha + phi*(urge1 - alpha) + depb + beta*(dep - depb),
  alpha ~ 1 + (1 | person),  # Latent intercept of urge
  depb ~ 1 + (1 | person),  # Latent mean of depression
  phi ~ 1 + (1 | person),  # Slope of latent mean-person centered lagged urge on urge
  beta ~ 1 + (1 | person),  # Slope of latent mean-person centered depression on urge
  nl = TRUE,
  center = FALSE # This seems key
) +
  gaussian()

#Set priors
p <- get_prior(model, data = d) %>% 
  mutate(
    prior = case_when(
      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
      TRUE ~ prior
    )
  )

fit2 <- brm(
  model,
  data = d,
  prior = p, 
  cores = MAX_CORES,
  threads = THREADS,
  iter = ITER,
  control = list(adapt_delta = 0.99),
  file = "models/m_mh5"
)

bind_rows(
  as_draws_df(fit1, variable = c("b_", "sd_", "sigma"), regex = TRUE),
  as_draws_df(fit2, variable = c("b_", "sd_", "sigma"), regex = TRUE) %>% 
    rename_with(~str_replace(., "depb", "depBI")), 
  .id = "Model"
) %>% 
  mutate(
    Model = factor(Model, labels = c("nlf()", "bf()")),
    ) %>% 
  select(-c(.chain, .iteration, .draw)) %>% 
  pivot_longer(-Model) %>% 
  ggplot(aes(value, name, col = Model, fill = Model)) +
  stat_halfeye(
    data = . %>% filter(Model == "nlf()"),
    side = "top", normalize = "xy", height = .5
  ) +
  stat_halfeye(
    data = . %>% filter(Model == "bf()"),
    side = "bottom", normalize = "xy", height = .5
  )
```


# Simulation study

```{r}
#| label: fig-me-in-means
#| eval: false
#| include: false
#| fig.cap: Illustration of the risks associated with using observed means for group-mean centering, and how those are avoided by using latent means to group-mean center instead.

# A suggestion for an illustrative figure, showing risks (bias) of observed mean centering, but also how latent mean centering avoids them.

tibble(
  n = seq(1, 100, by = .1),
  s = 1/sqrt(n)
  ) %>% 
  ggplot(aes(n, s)) +
  scale_y_continuous(
    "Measurement error in observed mean",
    limits = 0:1, 
    breaks = extended_breaks(),
    expand = expansion(c(.02, .05))
  ) +
  scale_x_continuous(
    "Number of measurements",
    expand = expansion(.025, .025)
  ) +
  # Theoretical measurement error in mean
  geom_line()
  # More lines or panels
  # Observed dynamic panel bias from a simulation
  # geom_line(something something)
  # Observed bias in other covariate from a simulation
  # geom_line(also something something)
  # Latent mean centering results--lines that are not biased
  # geom_line(...)
```


# Discussion

R [@rcoreteamLanguageEnvironmentStatistical2022]

\clearpage

# Acknowledgements

We would like to thank Mauricio Garnier-Villarreal, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424). Special thanks to Mauricio Garnier-Villarreal for being the first to post the syntax that the code here was based on. We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.

# References

::: {#refs custom-style="Bibliography"}
:::


<!-- Create an appendix with new page, table, and figure numbers. -->

\clearpage
\onecolumn

# Appendix A: Appendix McAppendix

<!-- Prepend table & figure numbers with S and reset counters -->
\setcounter{page}{1}
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r include-only-for-pdf}
#| include: !expr knitr::is_latex_output()
```
